An enhancer code compendium of early human neural and neural crest development

### 2025-05-12 manuscript.tex
Supplement

We generated human neural tube organoids to model early human neural and neural crest development. 
To this end we embedded single cells from two induced pluripotent stem cell (iPSC) lines (Sigma and BJ1) and one embryonic stem cell line (ESC; H9) into separate synthetic poly-ethylene glycol (PEG) based extracellular matrices. The organoids were incubated for 11 days and we supplemented the medium with retinoic acid (RA) and Smoothened agonist (SAG) from day 3 until day 5 (\textbf{Fig. 1a}). 
These conditions were previously shown to generate multiple neural progenitor and
neural crest states\cite{abdelfattahActuationEnhancesPatterning2021}. 
To characterize the organoid-derived cell types to a greater level of detail at both transcriptomic as well as chromatin accessibility level we performed 10x single-cell multiomics (combined scATAC-seq and
scRNA-seq) 
at four time points of differentiation (day 4-5, day 6-7, day 8-9 and day 10-11; \textbf{Fig. 1a}), 
resulting in 22,862 high quality cells (\textbf{Fig. 1b}). 
Note that similar transcriptomic states were obtained from all three cell lines (\textbf{Supplementary Fig. S1a-b}) demonstrating the robustness of the organoid system.\par

To evaluate the biological relevance of the cell types and states obtained from the organoid, 
we also performed 10x single cell multiomics on the head of a single four weeks post conception human embryo, 
resulting in 25,785 high quality cells (\textbf{Fig. 1c}). 
We annotated both the organoid and embryo cells based on the expression of marker genes
(\textbf{Supplementary Table S1}). 
In both systems we could identify neural progenitors, early differentiating neurons,
neural crest, facial mesenchyme and early differentiating peripheral neurons (\textbf{Fig. 1b-c}).
Additionally, in the organoid only we could identify a pluripotent stem cell population (\textbf{Fig. 1b}) 
while perivascular macrophages (PVM) / microglia (MIC) and otic placode cells were missing in the organoid system (\textbf{Fig. 1c}).\par

@@

Finally, to assess whether the organoid-derived cell types and states are also similar to the embryo at the level of chromatin accessibility
we identified sets of co-accessible regions in the organoid model using topic modeling 
and identified equivalent cells in the embryo where the same regions were also co-accessible.
We observed similar patterns of marker gene expression for the same cell state across the organoid and embryo (\textbf{Fig. 1i}) 
with an average Pearson correlation coefficient of 0.83 across matched states (\textbf{Supplementary Fig. S1c-d}).\par

@@

\begin{figure}[b!]
  \ContinuedFloat
  \caption*{\rule{\linewidth}{0.4pt}}
  \caption{
    (Continued)
    \textbf{d-f}, UMAP based on gene expression profiles of organoid (top) and embryo (bottom) cells types showing sub-clusters of neural progenitors (\textbf{d}), early differentiating neurons (\textbf{e}) and neural crest (\textbf{f}) along with expression of indicated genes in red-green-blue color scale.
    \textbf{g}, UMAP based on gene expression profiles of organoid (top; UMAP axis 1 and 2 shown) and embryo (bottom; UMAP
    axis 2 and 3 shown) for neural progenitor subcluster along with the expression of indicated genes in red-green-blue
    color scale. 
    \textbf{h}, UMAP based on gene expression profiles in organoid (top left) and embryo (top right) colored by
    cell cycle phase identity and quantification of percentage of each cell cycle phase stratified by cell
    type in the organoid (bottom left) and embryo (bottom right). 
    \textbf{i}, Average gene expression level in $log_{2}(CPM)$ (dot color) and fraction of cells expressing the gene (dot size) per cell state, defined by topic modeling, for organoid
    cells (top) and embryo cells (middle) and motif enrichment level as normalized enrichment score (NES; dot color) and
    number of regions enriched for the motif (dot size) per topic (x-axis) and motif (y-axis; bottom). PVM/MIC: perivascular
    microphage/microglia.
    Color scales for RGB scatter plots are as follows (in Log(CPM)). 
    For organoid: 
    Minimum is 0 and maximum is 1.5 for all genes except:
    IRX3, BAHRL1: [0, 1.0], 
    FOXG1, GAD1, UNXC, GATA2, SOX10: [0, 0.8] and 
    ZEB2 [0, 3.0], 
    For embryo: 
    Minimum is 0 and maximum is 1.5 for all genes except:
    FOXG1, GAD1, UNCX, GATA2, SOX10 [0, 0.8], 
    ISL1 [0, 1.0], 
    ZEB2 [0, 3.0] and
    TWIST1 [0, 2.0].
  }
\end{figure}

@@

We tested the performance of $DeepNeuralTube^o$ and $DeepNeuralTube^e$ to predict cell type-specific enhancer activity 
using data from the VISTA enhancer browser\cite{viselVISTAEnhancerBrowsera2007, kosickiVISTAEnhancerBrowser2025}.
Enhancers active in the neural tube (n=349) could be classified with an area under the receiver-operator curve
(auROC) of 0.75 and 0.80 using resp. $DeepNeuralTube^o$ and $DeepNeuralTube^e$ (\textbf{Fig. 2b}; and
area under the precision-recall curve (auPR) of 0.21 resp. 0.29 (\textbf{Supplementary Fig. S2a-b})). 
Enhancers active in the facial mesenchyme (n=143) could be classified with an auROC of 0.78 and 0.79 
using resp. $DeepNeuralTube^o$ and $DeepNeuralTube^e$ (\textbf{Fig. 2c}; 
and auPR of 0.17 for both models (\textbf{Supplementary Fig. S2a-b})). 
Of note, a fraction of neural tube enhancers (\SI{24}{\percent}) are predicted to be specific to pre-migratory neural crest by
both models. Given that the VISTA enhancers are annotated based on localization of enhancer activity only, and given
that it is difficult to visually distinguish between the neural tube and pre-migratory neural crest, this result 
is not surprising.\par

Next, we assessed the explainability of the models. Concretely we tested if the model learned known
TF binding motifs as features. To this end, we calculated pairwise similarities\cite{
  guptaQuantifyingSimilarityMotifs2007} between \textit{de novo} motifs learned by the model, 
and identified using TF-MoDISco\cite{shrikumarTechnicalNoteTranscription2018}, and known motifs that are enriched in cell type specific regions.
Clustering this pairwise similarity matrix resulted in 56 clusters (\textbf{Fig. 2d}). 
All clusters, except cluster 47 (organoid specific CG motif), contain known motifs enriched in regions specific to both 
organoid and embryo cell types (\textbf{Fig. 2e} and \textbf{Supplementary Table S2}).
More importantly, most clusters (\SI{70}{\percent}) contain at least one motif learned \textit{de novo} by the organoid and/or embryo model (\textbf{Fig. 2f-g} and \textbf{Supplementary Table S2}). 
Thus, $DeepNeuralTube^o$ and $DeepNeuralTube^e$ use known-motifs as features to classify regions based
on cell type-specific chromatin accessibility.\par

As an illustrative example, we investigated an enhancer located in one of the introns of \textit{GLI3} (hs111.0; \textbf{Supplementary Fig. S2c}). 
This region has neural tube specific activity in E11.5 mouse embryos (\textbf{Fig. 2h}). 
Both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ explain this region similarly (Pearson correlation of 0.79) 
and highlight the importance of multiple SOX TFBS (\textbf{Fig. 2h}). 
A version of this enhancer where 5\% of the nucleotides are changed (hs111.1) does not have neural tube specific enhancer activity and both models predict that this loss of activity is due to the loss of a single SOX TFBS (\textbf{Fig. 2h}).\par

Finally, to also investigate the performance of the models on migratory neural crest, 
a cell type for which there are no validated enhancers in the VISTA enhancer browser, 
we selected a genomic region that is specifically accessible in the migratory neural crest (\textbf{Supplementary Fig. S2d}). 
Both models predict this region to be specific to the migratory neural crest class (\textbf{Supplementary Fig. S2e})
and nucleotide importance scores reveal TFBS for a SOX monomer, a SOX dimer and AP2 (\textbf{Fig. 2i}). 
Next, we tested the enhancer activity of this region using an enhancer-reporter assay in chicken embryos. 
Indeed the enhancer was specifically active in migratory neural crest 24 hours after injection in an HH4 embryo (\textbf{Fig. 2i}).\par

@@

\begin{figure}[b!]
  \ContinuedFloat
  \caption*{\rule{\linewidth}{0.4pt}}
  \caption{
    (Continued) 
    \textbf{b}, receiver-operator curves for classifying neural tube (left) and facial mesenchyme enhancers using
    $DeepNeuralTube^o$ and $DeepNeuralTube^e$. 
    \textbf{c-e}, UMAP of TomTom\cite{guptaQuantifyingSimilarityMotifs2007} motif-to-motif similarities
    using both \textit{de novo} motifs from DeepNeuralTube and known enriched motifs. Each dot is a motif. Motifs are
    colored by: cluster identity (\textbf{c}), model of origin 
    (\textbf{d}; only showing \textit{de novo} motifs in this case) and whether the motif is found \textit{de novo} (Deep Learning) or by motif enrichment (cisTarget; \textbf{e}).
    \textbf{f}, Example representative motifs from clusters in b, showing motifs from motif enrichment (left) and found
    \textit{de novo} (right). 
    \textbf{g}, contribution score for reference and mutant version of VISTA enhancer hs111
    for $DeepNeuralTube^o$ (top two panels) and $DeepNeuralTube^e$ (bottom two panels) and enhancer-reporter
    assay for reference and mutant version of hs111 (right two panels)\cite{
    viselVISTAEnhancerBrowsera2007, kosickiVISTAEnhancerBrowser2025}. 
    \textbf{h}, contribution score for
    $DeepNeuralTube^o$ (top) and $DeepNeuralTube^e$ (bottom) of a genomic human neural crest
    enhancer (chr1:161200798-161201298; hg38) and enhancer-reporter assay in chicken embryo (HH4; right). Green is enhancer signal, pink is
    constitutive enhancer that serves as electroporation control.
  }
\end{figure}

@@

To investigate the enhancer code underlying dorso-ventral patterning of neural progenitors further, we clustered 
and aligned \textit{de novo} motifs, identified in dorso-ventral regions, based on pairwise motif similarities. 
Next, we calculated the average contribution for all identified motif instances per cluster and for each topic 
resulting in a dorso-ventral enhancer code-table (\textbf{Fig. 3f}).
Both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ found similar patterns underlying the dorso-ventral states and the 
importance of the instances of these patterns is highly correlated across both models (0.75 average Pearson correlation; \textbf{Fig. 3f}). 
For the most ventral state (organoid topic 33 and embryo topic 34) a combination of FOX, TEAD and RFX instances 
are important (\textbf{Fig. 3f-g}) and the importance of these instances correlates with the expression of respectively \textit{FOXA1/A2/P2/P4}, \textit{TEAD1} and \textit{RFX3/4} in both the organoid and embryo (\textbf{Fig. 3h} and Supplementary Fig. S3). 
In subsequent more dorsal neural progenitor domains we observe importance of NKX instances, correlating with \textit{NKX2-1/2} expression (\textbf{(Fig. 3f-h)}); 
PAX instances, correlating with \textit{PAX3/6} expression (\textbf{Fig. 3f-h} and Supplementary Fig. S3) and 
ZIC instances, correlating with \textit{ZIC1} expression (\textbf{Fig. 3f-h} and Supplementary Fig. S3). 
For all dorso-ventral progenitor domains, except the most ventral one (and the most dorsal one in the embryo), 
SOX instances are important correlating with the expression of \textit{SOX2/3} (\textbf{Fig. 3f-h} and Supplementary Fig. S3). 
Similarly, ZEB instances have a negative importance to the accessibility of regions in dorso-ventral domains except the most ventral one (\textbf{Fig. 3f} and Supplementary Fig. S3). 
Finally, we calculated the Jaccard index of genomic regions based on the presence of instances of each identified pattern to investigate which patterns often co-occur in the same regions (\textbf{Fig. 3i}). 
FOX, TEAD and RFX instances often co-occur in the same genomic region while instances of other patterns tend to be present in separate regions (\textbf{Fig. 3i}).\par

@@

It has been reported that FOXA1/2 is a pioneer factor\cite{iwafuchi-doiPioneerTranscriptionFactor2016, iwafuchiGeneNetworkTransitions2020}. 
Indeed, when centering on FOX instances we observe a strong Tn5 cut-site footprint (\textbf{Supplementary Fig. S4a}). 
We then asked if there is any difference between regions that are generally bound by FOXA2 in any cell state that expresses this TF and regions, bound by FOXA2, that are specific to the floorplate. 
For this purpose, we intersected FOXA2 ChIP-seq peaks in HepG2 and A549  with floorplate peaks containing FOX instances. 
We classified the regions that overlap with any of the ChIP-seq datasets as regions that are generally bound by FOXA2 and those that did not as floorplate specific. Quantifying both the number of FOX instances per region and 
the affinity of each individual instance (making use of the FOX motif score calculated using
FIMO\cite{grantFIMOScanningOccurrences2011, schreiberTangermemeBiologicalSequence2025}) reveals that floorplate 
specific FOXA2 target regions have fewer FOX instances compared to general FOXA2 target regions (\textbf{Supplementary Fig. S4b}). 
Furthermore, individual FOX instances in general FOXA2 target regions have higher affinity compared to FOX instances in
specific regions (\textbf{Supplementary Fig. S4c}). These results suggest that the overall affinity of general FOXA2 target regions 
is higher (either by having a greater number of binding sites or by having binding sites with higher affinity).

@@

\begin{figure}[b!]
  \ContinuedFloat
  \caption*{\rule{\linewidth}{0.4pt}}
  \caption{
    (Continued) 
    \textbf{b}, Boxplots of the expression of \textit{SHH} in organoid (top) and embryo (bottom) across different dorso-  ventral topics ($n_{topic 33}$ = 4,571, $n_{topic 38}$ = 4,521, $n_{topic 36}$ = 6,120, $n_{topic 54}$ = 2,808, and $n_{topic 48}$ = 5,390 for organoid cells;
      and $n_{topic 34}$ = 637, $n_{topic 38}$ = 1,093, $n_{topic 79}$ = 3,939, $n_{topic 88}$ = 1,861, and $n_{topic 58}$ = 1,773 for embryo cells).  With n representing the number of cells.
    \textbf{c}, Chromatin accessibility in organoid (top) and embryo (bottom) across dorso-ventral topics for locus chr7:155,799,664-155,827,483 ; hg38. SFPE2 and SFPE1 are highlighted. 
    \textbf{d}, Enhancer reporter assay in chicken embryo. Blue is DAPI, pink is constitutive electroportion control,
    green is GFP driven by SFPE1 enhancer.
    \textbf{e}, Nucleotide contribution scores for SFPE2 (left) and SFPE1 (right) for organoid topic 33 (top) and embryo topic 34 (bottom). 
    \textbf{f}, Code table showing average contribution score of pattern instances for each dorso-ventral topic. Contribution scores for $DeepNeuralTube^o$ are not outlined. Contribution scores for $DeepNeuralTube^e$ are outlined and negated. 
    \textbf{g}, Heatmap showing the maximum normalized contribution score for each of the top 1,000 regions per topic (y-axis) and pattern (x-axis) for dorso-ventral topics. 
    \textbf{h}, Heatmap showing scaled expression of transcription factors corresponding to identified patterns in organoid (left) and embryo (right) dorso-ventral topics. 
    \textbf{i}, Heatmap showing Jaccard index of regions based on the presence of instances of the identified patterns. 
    \textbf{j}, Nucleotide contribution score for chr18:77367159-77367659; hg38 for organoid topic 33 (top) and embryo
    topic 34 (bottom).
    \textbf{k}, Prediction scores for chr18:77367159-77367659; hg38 for doros-ventral topics in organoid (top) and
    embryo (bottom).
    \textbf{l}, Enhancer reporter assay in chicken embryo. Blue is DAPI, pink is constitutive electroportion control,
    green is GFP driven by chr18:77367159-77367659; hg38 enhancer. Scale bar represents \SI{500}{\micro\meter}
}
\end{figure}

@@

$DeepNeuralTube^e$ was able to distinguish the different AP domains but grouped topics of similar domains
% TODO: S.A.: How does the organoid model score all these AP topics?
together (i.e., Topic 59, 31 and 62; Topic 52 and 71), resulting in four distinct AP domains (\textbf{Fig.4 c-d}) and 
for this it made use of four distinct homeodomain TF binding motifs, one for each domain (\textbf{Fig. 4e}). 
From anterior to posterior this is \textit{TAATTA}, \textit{GGATTA}, \textit{TCATGWTGANTGA} and \textit{CATYMATCA}.
Notably, the patterns corresponding to the two posterior most topics correspond to homeobox dimer-motifs. 
Previously it had been hypothesized that genome-wide specific binding of homeodomain TFs is accomplished through
cooperative binding of two TFs\cite{mannChapter3Hox2009}. 
Indeed, the presence of instances of the \textit{TCATGWTGANTGA} motif correlates with the expression of both \textit{PAX5/8} and \textit{EN1/2} 
and the prescence of instances of the \textit{CATYMATCA} motif correlates with the expression of \textit{MEIS1/2}, \textit{PBX1/2} and \textit{HOXB3} (\textbf{Fig. 4d-e} and Supplementary Fig. S5). 
In contrast, patterns corresponding to the two anterior most topics correspond to monomer motifs, 
correlating with the expression of \textit{LHX2/9}, \textit{RAX} and \textit{SIX3/6} for the most anterior topic and 
with the expression of \textit{OTX1/2}, \textit{EMX2}, \textit{DMBX1} and \textit{LMX1A} for the second most anterior
topic (\textbf{Fig. 4d-e} and Supplementary Fig. S5)\par
%TODO: I'm surprised to see EMX2 (forebrain) together with OTX1/2 and DMBX1 (midbrain)

@@

with instances with 4 bp spacing being the most accessible (\textbf{Supplementary Fig. 10}). 

@@

@@

% TODO: see https://www.cell.com/information-for-authors/star-supplemental-information

@@

%\textsuperscript{b}More footnotes

>>>>

We generated human neural tube organoids to model early human neural and neural crest development. 
To this end we embedded single cells from two induced pluripotent stem cell (iPSC) lines (Sigma and BJ1) and one embryonic stem cell line (ESC; H9) into separate synthetic poly-ethylene glycol (PEG) based extracellular matrices. The organoids were incubated for 11 days and we supplemented the medium with retinoic acid (RA) and Smoothened agonist (SAG) from day 3 until day 5 (\textbf{Fig. 1a}). 
These conditions were previously shown to generate multiple neural progenitor and
neural crest states\cite{abdelfattahActuationEnhancesPatterning2021}. 
To characterize the organoid-derived cell types to a greater level of detail at both transcriptomic as well as chromatin accessibility level we performed 10x single-cell multiomics (combined scATAC-seq and
scRNA-seq) 
at four time points of differentiation (day 4-5, day 6-7, day 8-9 and day 10-11; \textbf{Fig. 1a}), 
resulting in 22,862 high quality cells (\textbf{Fig. 1b}). 
Note that similar transcriptomic states were obtained from all three cell lines (\textbf{Fig. S1a-b}) demonstrating the robustness of the organoid system.\par

To evaluate the biological relevance of the cell types and states obtained from the organoid, 
we also performed 10x single cell multiomics on the head of a single four weeks post conception human embryo, 
resulting in 25,785 high quality cells (\textbf{Fig. 1c}). 
We annotated both the organoid and embryo cells based on the expression of marker genes
(\textbf{Table S1}). 
In both systems we could identify neural progenitors, early differentiating neurons,
neural crest, facial mesenchyme and early differentiating peripheral neurons (\textbf{Fig. 1b-c}).
Additionally, in the organoid only we could identify a pluripotent stem cell population (\textbf{Fig. 1b}) 
while perivascular macrophages (PVM) / microglia (MIC) and otic placode cells were missing in the organoid system (\textbf{Fig. 1c}).\par

@@

Finally, to assess whether the organoid-derived cell types and states are also similar to the embryo at the level of chromatin accessibility
we identified sets of co-accessible regions in the organoid model using topic modeling 
and identified equivalent cells in the embryo where the same regions were also co-accessible.
We observed similar patterns of marker gene expression for the same cell state across the organoid and embryo (\textbf{Fig. 1i}) 
with an average Pearson correlation coefficient of 0.83 across matched states (\textbf{Fig. S1c-d}).\par

@@

\begin{figure}[b!]
  \ContinuedFloat
  \caption*{\rule{\linewidth}{0.4pt}}
  \caption{
    (Continued)
    \textbf{d-f}, UMAP based on gene expression profiles of organoid (top) and embryo (bottom) cells types showing sub-clusters of neural progenitors (\textbf{d}), early differentiating neurons (\textbf{e}) and neural crest (\textbf{f}) along with expression of indicated genes in red-green-blue color scale.
    \textbf{g}, UMAP based on gene expression profiles of organoid (top; UMAP axis 1 and 2 shown) and embryo (bottom; UMAP
    axis 2 and 3 shown) for neural progenitor subcluster along with the expression of indicated genes in red-green-blue
    color scale. 
    \textbf{h}, UMAP based on gene expression profiles in organoid (top left) and embryo (top right) colored by
    cell cycle phase identity and quantification of percentage of each cell cycle phase stratified by cell
    type in the organoid (bottom left) and embryo (bottom right). 
    \textbf{i}, Average gene expression level in $log_{2}(CPM)$ (dot color) and fraction of cells expressing the gene (dot size) per cell state, defined by topic modeling, for organoid
    cells (top) and embryo cells (middle) and motif enrichment level as normalized enrichment score (NES; dot color) and
    number of regions enriched for the motif (dot size) per topic (x-axis) and motif (y-axis; bottom). PVM/MIC: perivascular
    microphage/microglia.
    Color scales for RGB scatter plots are as follows (in Log(CPM)). 
    For organoid: 
    Minimum is 0 and maximum is 1.5 for all genes except:
    IRX3, BAHRL1: [0, 1.0], 
    FOXG1, GAD1, UNXC, GATA2, SOX10: [0, 0.8] and 
    ZEB2 [0, 3.0], 
    For embryo: 
    Minimum is 0 and maximum is 1.5 for all genes except:
    FOXG1, GAD1, UNCX, GATA2, SOX10 [0, 0.8], 
    ISL1 [0, 1.0], 
    ZEB2 [0, 3.0] and
    TWIST1 [0, 2.0].
    Also see Figure S1
  }
\end{figure}

@@

We tested the performance of $DeepNeuralTube^o$ and $DeepNeuralTube^e$ to predict cell type-specific enhancer activity 
using data from the VISTA enhancer browser\cite{viselVISTAEnhancerBrowsera2007, kosickiVISTAEnhancerBrowser2025}.
Enhancers active in the neural tube (n=349) could be classified with an area under the receiver-operator curve
(auROC) of 0.75 and 0.80 using resp. $DeepNeuralTube^o$ and $DeepNeuralTube^e$ (\textbf{Fig. 2b}; and
area under the precision-recall curve (auPR) of 0.21 resp. 0.29 (\textbf{Fig. S2a-b})). 
Enhancers active in the facial mesenchyme (n=143) could be classified with an auROC of 0.78 and 0.79 
using resp. $DeepNeuralTube^o$ and $DeepNeuralTube^e$ (\textbf{Fig. 2c}; 
and auPR of 0.17 for both models (\textbf{Fig. S2a-b})). 
Of note, a fraction of neural tube enhancers (\SI{24}{\percent}) are predicted to be specific to pre-migratory neural crest by
both models. Given that the VISTA enhancers are annotated based on localization of enhancer activity only, and given
that it is difficult to visually distinguish between the neural tube and pre-migratory neural crest, this result 
is not surprising.\par

Next, we assessed the explainability of the models. Concretely we tested if the model learned known
TF binding motifs as features. To this end, we calculated pairwise similarities\cite{
  guptaQuantifyingSimilarityMotifs2007} between \textit{de novo} motifs learned by the model, 
and identified using TF-MoDISco\cite{shrikumarTechnicalNoteTranscription2018}, and known motifs that are enriched in cell type specific regions.
Clustering this pairwise similarity matrix resulted in 56 clusters (\textbf{Fig. 2d}). 
All clusters, except cluster 47 (organoid specific CG motif), contain known motifs enriched in regions specific to both 
organoid and embryo cell types (\textbf{Fig. 2e} and \textbf{Table S2}).
More importantly, most clusters (\SI{70}{\percent}) contain at least one motif learned \textit{de novo} by the organoid and/or embryo model (\textbf{Fig. 2f-g} and \textbf{Table S2}). 
Thus, $DeepNeuralTube^o$ and $DeepNeuralTube^e$ use known-motifs as features to classify regions based
on cell type-specific chromatin accessibility.\par

As an illustrative example, we investigated an enhancer located in one of the introns of \textit{GLI3} (hs111.0; \textbf{Fig. S2c}). 
This region has neural tube specific activity in E11.5 mouse embryos (\textbf{Fig. 2h}). 
Both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ explain this region similarly (Pearson correlation of 0.79) 
and highlight the importance of multiple SOX TFBS (\textbf{Fig. 2h}). 
A version of this enhancer where 5\% of the nucleotides are changed (hs111.1) does not have neural tube specific enhancer activity and both models predict that this loss of activity is due to the loss of a single SOX TFBS (\textbf{Fig. 2h}).\par

Finally, to also investigate the performance of the models on migratory neural crest, 
a cell type for which there are no validated enhancers in the VISTA enhancer browser, 
we selected a genomic region that is specifically accessible in the migratory neural crest (\textbf{Fig. S2d}). 
Both models predict this region to be specific to the migratory neural crest class (\textbf{Fig. S2e})
and nucleotide importance scores reveal TFBS for a SOX monomer, a SOX dimer and AP2 (\textbf{Fig. 2i}). 
Next, we tested the enhancer activity of this region using an enhancer-reporter assay in chicken embryos. 
Indeed the enhancer was specifically active in migratory neural crest 24 hours after injection in an HH4 embryo (\textbf{Fig. 2i}).\par

@@

\begin{figure}[b!]
  \ContinuedFloat
  \caption*{\rule{\linewidth}{0.4pt}}
  \caption{
    (Continued) 
    \textbf{b}, receiver-operator curves for classifying neural tube (left) and facial mesenchyme enhancers using
    $DeepNeuralTube^o$ and $DeepNeuralTube^e$. 
    \textbf{c-e}, UMAP of TomTom\cite{guptaQuantifyingSimilarityMotifs2007} motif-to-motif similarities
    using both \textit{de novo} motifs from DeepNeuralTube and known enriched motifs. Each dot is a motif. Motifs are
    colored by: cluster identity (\textbf{c}), model of origin 
    (\textbf{d}; only showing \textit{de novo} motifs in this case) and whether the motif is found \textit{de novo} (Deep Learning) or by motif enrichment (cisTarget; \textbf{e}).
    \textbf{f}, Example representative motifs from clusters in b, showing motifs from motif enrichment (left) and found
    \textit{de novo} (right). 
    \textbf{g}, contribution score for reference and mutant version of VISTA enhancer hs111
    for $DeepNeuralTube^o$ (top two panels) and $DeepNeuralTube^e$ (bottom two panels) and enhancer-reporter
    assay for reference and mutant version of hs111 (right two panels)\cite{
    viselVISTAEnhancerBrowsera2007, kosickiVISTAEnhancerBrowser2025}. 
    \textbf{h}, contribution score for
    $DeepNeuralTube^o$ (top) and $DeepNeuralTube^e$ (bottom) of a genomic human neural crest
    enhancer (chr1:161200798-161201298; hg38) and enhancer-reporter assay in chicken embryo (HH4; right). Green is enhancer signal, pink is
    constitutive enhancer that serves as electroporation control.
    Also see Figure S2
  }
\end{figure}

@@

To investigate the enhancer code underlying dorso-ventral patterning of neural progenitors further, we clustered 
and aligned \textit{de novo} motifs, identified in dorso-ventral regions, based on pairwise motif similarities. 
Next, we calculated the average contribution for all identified motif instances per cluster and for each topic 
resulting in a dorso-ventral enhancer code-table (\textbf{Fig. 3f}).
Both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ found similar patterns underlying the dorso-ventral states and the 
importance of the instances of these patterns is highly correlated across both models (0.75 average Pearson correlation; \textbf{Fig. 3f}). 
For the most ventral state (organoid topic 33 and embryo topic 34) a combination of FOX, TEAD and RFX instances 
are important (\textbf{Fig. 3f-g}) and the importance of these instances correlates with the expression of respectively \textit{FOXA1/A2/P2/P4}, \textit{TEAD1} and \textit{RFX3/4} in both the organoid and embryo (\textbf{Fig. 3h} and Fig. S3). 
In subsequent more dorsal neural progenitor domains we observe importance of NKX instances, correlating with \textit{NKX2-1/2} expression (\textbf{(Fig. 3f-h)}); 
PAX instances, correlating with \textit{PAX3/6} expression (\textbf{Fig. 3f-h} and Fig. S3) and 
ZIC instances, correlating with \textit{ZIC1} expression (\textbf{Fig. 3f-h} and Fig. S3). 
For all dorso-ventral progenitor domains, except the most ventral one (and the most dorsal one in the embryo), 
SOX instances are important correlating with the expression of \textit{SOX2/3} (\textbf{Fig. 3f-h} and Fig. S3). 
Similarly, ZEB instances have a negative importance to the accessibility of regions in dorso-ventral domains except the most ventral one (\textbf{Fig. 3f} and Fig. S3). 
Finally, we calculated the Jaccard index of genomic regions based on the presence of instances of each identified pattern to investigate which patterns often co-occur in the same regions (\textbf{Fig. 3i}). 
FOX, TEAD and RFX instances often co-occur in the same genomic region while instances of other patterns tend to be present in separate regions (\textbf{Fig. 3i}).\par

@@

It has been reported that FOXA1/2 is a pioneer factor\cite{iwafuchi-doiPioneerTranscriptionFactor2016, iwafuchiGeneNetworkTransitions2020}. 
Indeed, when centering on FOX instances we observe a strong Tn5 cut-site footprint (\textbf{Fig. S4a}). 
We then asked if there is any difference between regions that are generally bound by FOXA2 in any cell state that expresses this TF and regions, bound by FOXA2, that are specific to the floorplate. 
For this purpose, we intersected FOXA2 ChIP-seq peaks in HepG2 and A549  with floorplate peaks containing FOX instances. 
We classified the regions that overlap with any of the ChIP-seq datasets as regions that are generally bound by FOXA2 and those that did not as floorplate specific. Quantifying both the number of FOX instances per region and 
the affinity of each individual instance (making use of the FOX motif score calculated using
FIMO\cite{grantFIMOScanningOccurrences2011, schreiberTangermemeBiologicalSequence2025}) reveals that floorplate 
specific FOXA2 target regions have fewer FOX instances compared to general FOXA2 target regions (\textbf{Fig. S4b}). 
Furthermore, individual FOX instances in general FOXA2 target regions have higher affinity compared to FOX instances in
specific regions (\textbf{Fig. S4c}). These results suggest that the overall affinity of general FOXA2 target regions 
is higher (either by having a greater number of binding sites or by having binding sites with higher affinity).

@@

\begin{figure}[b!]
  \ContinuedFloat
  \caption*{\rule{\linewidth}{0.4pt}}
  \caption{
    (Continued) 
    \textbf{b}, Boxplots of the expression of \textit{SHH} in organoid (top) and embryo (bottom) across different dorso-  ventral topics ($n_{topic 33}$ = 4,571, $n_{topic 38}$ = 4,521, $n_{topic 36}$ = 6,120, $n_{topic 54}$ = 2,808, and $n_{topic 48}$ = 5,390 for organoid cells;
      and $n_{topic 34}$ = 637, $n_{topic 38}$ = 1,093, $n_{topic 79}$ = 3,939, $n_{topic 88}$ = 1,861, and $n_{topic 58}$ = 1,773 for embryo cells).  With n representing the number of cells.
    \textbf{c}, Chromatin accessibility in organoid (top) and embryo (bottom) across dorso-ventral topics for locus chr7:155,799,664-155,827,483 ; hg38. SFPE2 and SFPE1 are highlighted. 
    \textbf{d}, Enhancer reporter assay in chicken embryo. Blue is DAPI, pink is constitutive electroportion control,
    green is GFP driven by SFPE1 enhancer.
    \textbf{e}, Nucleotide contribution scores for SFPE2 (left) and SFPE1 (right) for organoid topic 33 (top) and embryo topic 34 (bottom). 
    \textbf{f}, Code table showing average contribution score of pattern instances for each dorso-ventral topic. Contribution scores for $DeepNeuralTube^o$ are not outlined. Contribution scores for $DeepNeuralTube^e$ are outlined and negated. 
    \textbf{g}, Heatmap showing the maximum normalized contribution score for each of the top 1,000 regions per topic (y-axis) and pattern (x-axis) for dorso-ventral topics. 
    \textbf{h}, Heatmap showing scaled expression of transcription factors corresponding to identified patterns in organoid (left) and embryo (right) dorso-ventral topics. 
    \textbf{i}, Heatmap showing Jaccard index of regions based on the presence of instances of the identified patterns. 
    \textbf{j}, Nucleotide contribution score for chr18:77367159-77367659; hg38 for organoid topic 33 (top) and embryo
    topic 34 (bottom).
    \textbf{k}, Prediction scores for chr18:77367159-77367659; hg38 for doros-ventral topics in organoid (top) and
    embryo (bottom).
    \textbf{l}, Enhancer reporter assay in chicken embryo. Blue is DAPI, pink is constitutive electroportion control,
    green is GFP driven by chr18:77367159-77367659; hg38 enhancer. Scale bar represents \SI{500}{\micro\meter}. 
    Also see Figure S3 and S4.
}
\end{figure}

@@

$DeepNeuralTube^e$ was able to distinguish the different AP domains but grouped topics of similar domains
% TODO: S.A.: How does the organoid model score all these AP topics?
together (i.e., Topic 59, 31 and 62; Topic 52 and 71), resulting in four distinct AP domains (\textbf{Fig.4 c-d}) and 
for this it made use of four distinct homeodomain TF binding motifs, one for each domain (\textbf{Fig. 4e}). 
From anterior to posterior this is \textit{TAATTA}, \textit{GGATTA}, \textit{TCATGWTGANTGA} and \textit{CATYMATCA}.
Notably, the patterns corresponding to the two posterior most topics correspond to homeobox dimer-motifs. 
Previously it had been hypothesized that genome-wide specific binding of homeodomain TFs is accomplished through
cooperative binding of two TFs\cite{mannChapter3Hox2009}. 
Indeed, the presence of instances of the \textit{TCATGWTGANTGA} motif correlates with the expression of both \textit{PAX5/8} and \textit{EN1/2} 
and the prescence of instances of the \textit{CATYMATCA} motif correlates with the expression of \textit{MEIS1/2}, \textit{PBX1/2} and \textit{HOXB3} (\textbf{Fig. 4d-e} and Fig. S5). 
In contrast, patterns corresponding to the two anterior most topics correspond to monomer motifs, 
correlating with the expression of \textit{LHX2/9}, \textit{RAX} and \textit{SIX3/6} for the most anterior topic and 
with the expression of \textit{OTX1/2}, \textit{EMX2}, \textit{DMBX1} and \textit{LMX1A} for the second most anterior
topic (\textbf{Fig. 4d-e} and Fig. S5)\par
%TODO: I'm surprised to see EMX2 (forebrain) together with OTX1/2 and DMBX1 (midbrain)

@@

with instances with 4 bp spacing being the most accessible (\textbf{Fig. 10}). 

@@

The authors declare no competing interests.

@@

% see https://www.cell.com/information-for-authors/star-supplemental-information

\begin{description}
  \item Document S1. Table S1
  \item Table S2. Excel file containing motifs for clustering in Figure 2.
\end{description}

@@

%\textsuperscript{b}More footnotes
%
%NOTE: For the journal Cell, supplemental figures should instead be uploaded as separate files, the same process as the main figures; their titles and legends should be placed in the main text, directly after the main figure legends and any tables, under a “supplemental figures” heading
